\chapter*{Acknowledgments}
\addcontentsline{toc}{chapter}{Acknowledgements}
\begin{noparindent}
\SingleSpacing

I could not have written this thesis without the help of a great many people, and I would like to acknowledge a few of them here.

Jake Sullivan, Robert Booth, and John Farquhar spoke with me as part of my research, and Dr. Farquhar was kind enough to mail me his book, thesis, and a copy of one of his articles.

A huge thank you to Professor Lind for agreeing to serve as my second reader, even though, at the time of this writing, I still have not met you! I look forward to changing that.

Professor Valentino went above and beyond his role as my steward through the thesis program, offering advice, corrections, and thoughts that significantly improved the quality of my argument. If you don't like the title---blame him.

I wish I could name all the friends that helped me get to where I am today, but I would probably forget someone and then I would feel bad. They know who they are. Without their support, advice, and constant reassurance that they would remain my friend even though I ``can't hang,'' I would never have made it past the winter.

There is no person at Dartmouth to whom I owe more for my academic development than Professor William Wohlforth. In the years since he wrote ``wait a min, is this guy serious'' in the comments for my final paper, he has consistently demonstrated a willingness to devote his time to energy towards my intellectual development. His deeply considered thoughts on my work are one of the main reasons it is (hopefully) good enough for someone else to take seriously.

At the times during this project when I struggled the most, my parents did what they have done for me my whole life---give me something new to read. My mom found the book that inspired my conclusions and my dad found the quotes that prove them to be true. Their influence can be found in the copy-editing, the citations, the theory, the evidence, and the love of learning that inspired me to write this thesis in the first place.

Mom, Dad, Sophia, Andrew---thank you. I love you very much.

---Alex

\end{noparindent}
\newpage
